Diversity and evolution of centromere repeats in the maize genome

### 2014-04-30 centc.tex
intro done

Centromere repeats are found in most eukaryotes and play a critical role in kinetochore formation.  Though they exhibit considerable diversity both within and among species, little is understood about the mechanisms that drive centromere repeat evolution.  Here, we use maize as a model to investigate how a complex history involving polyploidy, fractionation, and recent domestication has impacted the diversity of the maize CentC repeat.  We first validate the existence of long tandem arrays of repeats in maize and other taxa in \emph{Zea}.  We show that genetic similarity among CentC copies is highest within these arrays, suggesting that tandem duplications are the primary mechanism for the generation of new copies.  In spite of this, we see little evidence that CentC variants form distinct genetic groups, instead finding that homoplasious mutations have likely homogenized CentC diversity.  We identify Cent4, a repeat initially thought to be specific to the chromosome 4 centromere, as an LTR retrotransposon that has increased drastically in frequency through domestication.  Although the two ancestral subgenomes of maize have contributed nearly equal numbers of centromeres, our analysis shows that a vast majority of all CentC repeats derive from only one of the parental genomes.  Finally, by comparing maize with its wild progenitor teosinte, we find that the abundance of CentC has decreased through domestication.

\keywords{Centromere \and Evolution \and Subgenome \and sweetscience}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

@@

Maize centromeres are comprised primarily of the 156bp satellite repeat CentC and a family of retrotransposons (CRM), both of which interact with  kinetochore proteins such as CENH3 \citep{Wolfgruber2009, Zhong2002}, and both repeats show considerable variation in local abundance across taxa \citep{Albert2010}.  But while there is considerable effort in investigating the molecular function of maize centromere repeats \citep{Ananiev1998B, Nagaki2003, Wolfgruber2009}, we know comparatively little about the evolution of the sequences themselves. CRM elements are better understood, including the age and insertion preferences of different CRM families \citep{Wolfgruber2009, Sharma2008}.  In contrast,  studies \citep{Wang2012} have only examined the flanking sequences to CentC islands despite their known association with functional centromeres.  To date, there is no in-depth characterization of the genetic diversity of centromere repeats in the maize genome.

In this paper, we describe the patterns of diversity of centromere repeats across the maize genome.  We investigate the genetic diversity of these repeats, including whether the differential ancestry of maize centromeres \citep{Wang2012} has led to chromosome-specific variation as found in other species \citep{Kawabe2005, Pontes2004} and how genetic relatedness among individual repeats varies spatially along chromosomes and across the genome.  We then compare centromere abundance across a number of maize lines, including to its wild relatives, the teosintes.  We find that CentC copies do not form genetic groups consistent with ancient whole genome duplications or chromosome specificity.  Instead, we show higher genetic similarity within clusters, potentially indicating the predominance of tandem duplications in the formation of new CentC copies.  Lastly, we use low coverage sequencing and cytological analyses data to show that domesticated maize has less CentC than its wild relatives.

>>>>

Centromere repeats are found in most eukaryotes and play a critical role in kinetochore formation.  Though they exhibit considerable diversity both within and among species, little is understood about the mechanisms that drive centromere repeat evolution.  Here, we use maize as a model to investigate how a complex history involving polyploidy, fractionation, and recent domestication has impacted the diversity of the maize CentC repeat.  We first validate the existence of long tandem arrays of repeats in maize and other taxa in the genus \emph{Zea}.  We find that genetic similarity among CentC copies is highest within these arrays, suggesting that tandem duplications are the primary mechanism for the generation of new copies.  In spite of this, we see little evidence that CentC variants form distinct genetic groups, instead finding that homoplasious mutations have likely homogenized CentC diversity.   Although the two ancestral subgenomes of maize have contributed nearly equal numbers of centromeres, our analysis shows that a vast majority of all CentC repeats derive from only one of the parental genomes.  Finally, by comparing maize with its wild progenitor teosinte, we find that the abundance of CentC, but not the pericentromeric repeat Cent4, has decreased through domestication.

\keywords{Centromere \and Evolution \and Subgenome}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

@@

Maize centromeres are comprised primarily of the 156bp satellite repeat CentC and the CRM family of retrotransposons.  Both repeats interact with  kinetochore proteins such as CENH3 \citep{Wolfgruber2009, Zhong2002} and show  variation in  abundance across taxa \citep{Albert2010}.  But while there is considerable effort in investigating the molecular function of maize centromere repeats \citep{Ananiev1998B, Nagaki2003, Wolfgruber2009}, we know comparatively little about the evolution of the sequences themselves. CRM elements are better understood, including the age and insertion preferences of different CRM families \citep{Wolfgruber2009, Sharma2008}.  In contrast, there is no in-depth characterization of the genetic diversity of centromere repeats in the maize genome.  
%  studies \citep{Wang2012} have only examined the flanking sequences to CentC islands despite their known association with functional centromeres.  To date,

In this paper, we describe the patterns of diversity of centromere repeats across the maize genome.  We investigate the genetic diversity of these repeats, including whether the differential ancestry of maize centromeres \citep{Wang2012} has led to chromosome-specific variation similar to that seen in other species \citep{Kawabe2005, Pontes2004} and how genetic relatedness among individual repeats varies spatially across the genome.  We then compare centromere abundance across a number of samples of maize and its wild relative teosinte.  We find that CentC copies do not form genetic groups consistent with ancient whole genome duplications or chromosome specificity, but instead that most of the large blocks of CentC appear to have been inherited from only one of the ancestral subgenomes of maize.  We show higher genetic similarity of CentC repeats within clusters, indicating the predominance of tandem duplications in the formation of new CentC copies.  Lastly, we use low coverage sequencing and cytological analyses data to show that domesticated maize has less CentC than its wild relatives.
